Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-15 analysis/paper/paper.qmd
finish editing methods - archaeological sites

After the first evidence of the Korean Late Paleolithic discovered at Seokjangri site including blade and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we gathered those that retained the widest and thickest part of specimens. Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but accompanied with other Late Paleoltihic artifacts. Those criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map). The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of Korean local museums and archaeological institutions.

We distinguished multiple assemblages in one site where numerous excavations have been conducted in different spots and by different institutions under the same site name. For example, Suyanggae site, one of the Korean National Heritage sites, has been excavated more than ten times since 1980 by the local university museum and later by archaeology research institutions. There are six different excavation spots apart from each other from a few meters to a few kilometers. The archaeologists identified multiple assemblages in each excavation or even a trench by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits. In this research, we used four assemblages from the Suyanggae site and we separated one assemblage among the four by using a different site name, Hajinri. Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka). According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al._2018]. We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time. We consider Bonggok site, which has the second-earliest dates (41.5 ka) in our collection, is the actual earliest site. Bonggok site includes blades or elongate flakes but it is hard to define without accompanying blade core [@park2022did].

>>>>

Since the first evidence of the Korean Late Paleolithic was discovered at Seokjangri site, including blades and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were]. Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that retained their widest and thickest parts. Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Palaeoltihic artifacts. Applying these sampling criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map). The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.

We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name. For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions. There are six different excavation locations that range from a few meters apart to a few kilometers. Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits. As a result, in this research we used four assemblages from the Suyanggae site. We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location. Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka). According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al._2018]. We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are so early and also so highly standardized and refined, which are only found at other sites much later in time. We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points. Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].
